Hypertension and microbiome

### 2018-12-31 rrbiome.Rmd
better subgrouping

Phenotypical data between genus and species all.equal `r
isTRUE(all.equal(dset.genus, dset.species))`. Dimension for data
frames are `r dim(dset.genus)` and `r dim(dset.species)`.

@@

# Alpha diversity

Dset with alpha diversity

```{r alpha diversity}

alphadiversity <- microbiome::global(pseq_filtered, index="shannon")
dset.alpha <- merge(dset, alphadiversity, by=0, all=TRUE)

```

```{r alpha diversity scatterplot, fig.width=10, fig.height=4}

ggplot((dset.alpha %>%
        select(BL_AGE, BMI, SYSTM, DIASM, PULSEPRESSURE, MAP, SEX, diversities_shannon) %>%
        gather(Covariate, Value, -SEX, -diversities_shannon)), 
       aes(x=diversities_shannon, y=Value)) +
    facet_wrap(~Covariate, ncol=4, scales = "free") +
    geom_point(size=0.1, shape=1, aes(colour = SEX)) +
    geom_smooth(method=lm, se=TRUE, size=0.5, color = 'Black') +
    stat_poly_eq(formula = "y ~ x",
                 aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")),
                 label.y = 0,
                 parse = TRUE,
                 size = 3) +   
    ylim(0, 200) +
    ggplot2::theme_minimal()

```

Linear models

```{r lm model}
lm(MAP ~ diversities_shannon, data=dset.alpha) %>%
    summary(.) %>% coef %>% kable(., digits=4)
lm(MAP ~ BL_AGE + SEX + diversities_shannon, data=dset.alpha) %>%
    summary(.) %>% coef %>% kable(., digits=4)
lm(MAP ~ BL_AGE + SEX + BMI + CURR_SMOKE + Q57X + BP_TREAT + diversities_shannon, data=dset.alpha) %>%
    summary(.) %>% coef %>% kable(., digits=4)

```

# Beta diversity

Beetadiversiteetti-matriisin lasku. Tiedosista comp_bc_dissimilarity_matrix.R ja käyttö comp_adonis_covariates.R.

```{r vegdist}

phf_profile.rel <- microbiome::transform(pseq_filtered, 'compositional')
otu <- microbiome::abundances(phf_profile.rel)

if (!file.exists("rds/bray.dist.m.rds")) {
    bray.dist.m <- vegdist(t(otu), method="bray")
    saveRDS(bray.dist.m, file = "rds/bray.dist.m.rds")
} else {
    bray.dist.m <- readRDS("rds/bray.dist.m.rds")
}

dist.matrix <- as.matrix(bray.dist.m)
attr(dist.matrix, "method") <- "bray"

```

```{r beta hclust, fig.width = 5, fig.height = 4, dpi = 300}
gplots::heatmap.2(dist.matrix,
                  distfun = dist,
                  trace = "none",
                  hclustfun = hclust,
                  key.title = NA,
                  dendrogram = "column",
                  density.info = "none",
                  labRow = NA,
                  labCol = NA)
```

Mallin lasku Adoniksella

```{r adonis1}
adonis(dist.matrix ~ MAP, 
       data = meta(pseq_filtered),
       permutations = 99)$aov.tab %>%
                            as.data.frame %>%
                            kable
```

```{r adonis2}
adonis(dist.matrix ~ BL_AGE + SEX + MAP,
       data = meta(pseq_filtered),
       permutations = 99)$aov.tab %>%
                            as.data.frame %>%
                            kable
```

```{r adonis3}
adonis(dist.matrix ~ MAP + BL_AGE + SEX + BMI + CURR_SMOKE + Q57X + BP_TREAT,
       data = meta(pseq_filtered),
       permutations = 99)$aov.tab %>%
                            as.data.frame %>%
                            kable
```

# PCA

```{r pca}
pseq.clr <- microbiome::transform(pseq_filtered, "clr")
X <- as.matrix(t(abundances(pseq.clr)))
prcomp <- stats::prcomp(X)
```

```{r pca labels}
percentage <- round(prcomp$sdev/sum(prcomp$sdev) * 100, 2)
pca.labels <- paste0(colnames(prcomp$x), " (", paste0(as.character(percentage), "%", ")"))
```

```{r PCA plots categorical, fig.width = 10, fig.height = 10, dpi = 300}
ggplot((cbind(meta(pseq_filtered), prcomp$x[,1:2]) %>%
        select(PC1, PC2, BP_TREAT, CURR_SMOKE, DIAB, BL_USE_RX_C09, Q57X,
               BL_USE_RX_C03, BL_USE_RX_C07, BL_USE_RX_C08, HYPERTENSION, SEX) %>%
        gather(Covariate, Value, -PC1, -PC2)), 
       aes(x=PC1, y=PC2)) +
    facet_wrap(~Covariate, ncol=2, scales = "free") +
    geom_point(aes(color = Value), size = 0.5) +
    scale_color_brewer(palette="Dark2") +
    xlab(pca.labels[1]) +
    ylab(pca.labels[2]) +
    ggplot2::theme_minimal()

```

```{r PCA plots continuous, fig.width = 10, fig.height = 10, dpi = 300}

ggplot((cbind(meta(pseq_filtered), prcomp$x[,1:2]) %>%
        select(PC1, PC2, BL_AGE, SYSTM, DIASM, BMI, MAP, PULSEPRESSURE) %>%
        gather(Covariate, Value, -PC1, -PC2)), 
       aes(x=PC1, y=PC2)) +
    facet_wrap(~Covariate, ncol=2, scales = "free") +
    geom_point(aes(color = Value), size = 0.5) +
    scale_color_gradientn(colours = rainbow(12)) +
    xlab(pca.labels[1]) +
    ylab(pca.labels[2]) +
    ggplot2::theme_minimal()

```

# Genus

Aaron work-december-2018 prepare_maaslin_input.R

```{r prepare maaslin}

pheno <- t(meta(sample_data(pseq_filtered)))
sample.ids <- colnames(pheno)
phfg.pseq.rel <- microbiome::transform(pseq_filtered, "compositional")
abud.rel <- abundances(phfg.pseq.rel)

rownames(abud.rel) <- sapply(rownames(abud.rel),
                             function (genus) gsub("(.*) \\(+(.+)\\)", '\\1.\\2', genus))

merged <- rbind(sample.ids, pheno, abud.rel)
write.table(t(merged), file="merged_phfinrisk_genus.tsv", sep='\t', 
            quote=FALSE, row.names=FALSE, col.names=TRUE)

variables <- rownames(pheno)
variables_wo_ctrls <- variables[!variables %in% c("BL_AGE","BMI","SEX")]
data <- file("maaslin_config.config" , open='wt')
pcl.contents <- c("Matrix: Metadata",
                  paste(c("Read_PCL_Rows: BL_AGE,BMI,SEX",variables_wo_ctrls), collapse=","),
                  "",
                  "Matrix: Adundance",
                  "Read_PCL_Rows: Archaea.Archaea-")
writeLines(pcl.contents, con=data)
close(data)
```

```{r run maaslin, eval = FALSE} 
Maaslin("merged_phfinrisk_genus.tsv",
        "maaslin",
        strInputConfig = "maaslin_config.config",
        strModelSelection="none",
        dSignificanceLevel = 1.0, 
        fAllvAll = TRUE,
        strForcedPredictors = c("BL_AGE", "BMI", "SEX"))

```

Data framen luku *comp_maaslin_results_from_single_job.R* ja plottaus käyttäen *comp_maaslin_plots.R* ja *edgeR_plot_functions.R*.

```{r Reading maaslins output}

maas.results.df <- read.table("maaslin/merged_phfinrisk_genus.txt", header=TRUE, sep='\t')
maas.results.df[,'Covariate'] <- maas.results.df[,'Variable']
maas.results.df <- merge(maas.results.df,
                         names.dset[,c('Covariate','Category', 'Name')],
                         by=c('Covariate'))

maas.results.df[,'Maaslin.Q.value'] <- maas.results.df[,'Q.value']
maas.results.df[,'Qval'] <- p.adjust(maas.results.df[,'P.value'], method="BH")
maas.results.df[,'beta'] <- maas.results.df[,'Coefficient']
maas.results.df.signf <- dplyr::filter(maas.results.df, Qval < 0.05)

```

```{r scatter plots}

sysm.signf.features <- maas.results.df.signf %>%
    dplyr::filter(Covariate == "SYSTM") %>%
    pull(Feature) %>%
    as.character() %>%
    gsub("_", ".", .)
df.merged <- t(merged) %>% as.data.frame
```

```{r maaslin scatter plot, fig.width = 10, fig.height = 20, dpi = 300}

#df.merged %>% pull("Rikenella.Bacteria") %>% as.character I%>% as.numeric %>% sqrt %>% asin

ggplot(df.merged %>%
       select(sysm.signf.features, SYSTM, SEX) %>%
       gather(Covariate, Value, sysm.signf.features) %>%
       mutate(SYSTM = as.numeric(as.character(SYSTM)),
              Value = as.numeric(as.character(Value))),
       aes(x = asin(sqrt(Value)), y = SYSTM)) +
    facet_wrap(~Covariate, ncol=2, scales = "free") +
    geom_point(aes(color = SEX), size = 0.5) +
    scale_y_continuous(breaks=seq(0,200,40)) +
    geom_smooth(method=lm, se=TRUE, size=0.5, color = 'Black') +
    stat_poly_eq(formula = "y ~ x",
                 aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")),
                 label.y = 0,
                 parse = TRUE,
                 size = 3) +   
    ggplot2::theme_minimal()

```

```{r maaslin plots, fig.width = 10, fig.height = 4, dpi = 300}
source("edgeR_plot_functions.R")

plot_maaslin_heatmap_final(maas.results.df.signf,
                      xlab.text="Genera",
                      beta.maxabs=max(abs(maas.results.df.signf[,'beta'])))

```

```{r scp html}
system('scp rrbiome.html utu:articletwo')
```

>>>>

Phenotypical data between genus and species all.equal `r isTRUE(all.equal(dset.genus, dset.species))`. Dimension for data frames are (`r dim(dset.genus)`) and (`r dim(dset.species)`).

@@

```{r scp html, cache = FALSE}
system('scp rrbiome.html utu:articletwo')
```
